---
title: "Examples"
---

---
class: center, middle, inverse

# Examples

---
# Simulated data

- Level change $(t=40)$ and growth change $(t=80)$.

```{r sim-data-viz-2}
ggplot(data = data_sim, aes(x = t, y = y)) +
  geom_point(size = 2) +
  geom_vline(xintercept = c(40, 80), linetype = "dotted", col = "red",
             size = 1.4) +
  scale_y_continuous(breaks = scales::pretty_breaks(8)) +
  scale_x_continuous(breaks = scales::pretty_breaks(10)) +
  labs(x = "Time", y = "y")
```

---
# The linear growth level model 

The process equations are
$$
\begin{eqnarray}
Y_t &=& \mu_t + \nu_t, \\
\mu_t &=& \mu_{t-1} + \beta_{t} + \partial\mu_t, \\ 
\beta_t &=& \beta_{t-1} + \partial\beta_t
\end{eqnarray}
$$
where

- $\mu_t$ is the level of the process at time $t$.

- $\beta_t$ is the incremental growth in the level of the series over time $t-1$ to $t$.

### Discount factors:

- $\delta_T = (0.97, 0.97) \rightarrow \delta^{\prime}_T = (0.10, 0.20)$.

---
# Using the `Monitoring` class

```{python fit-monitor-simulated,  echo=TRUE}
from pybats.dglm import dlm
from pybats_detection.monitor import Monitoring
a = np.array([100, 0])
R = np.eye(2)
R[0, 0] = 100
R[1, 1] = 100
mod = dlm(a, R, ntrend=2, deltrend=0.97)
monitor = Monitoring(mod=mod)
monitor_results = monitor.fit(
    y=r.data_sim["y"],
    h=4, tau=0.135,
    discount_factors={"trend": [0.10, 0.20]},
    distr_type="location",
    bilateral=True,
    prior_length=10)
```

```{python fit-no-monitor-simulated, include=FALSE}
no_monitor_results = monitor.fit(
    y=r.data_sim["y"],
    h=10000, tau=0.135,
    discount_factors={"trend": [0.20, 0.8]},
    distr_type="location",
    bilateral=True,
    prior_length=4)

# Predictive
df_monitor = monitor_results["filter"]["predictive"].copy()
df_no_monitor = no_monitor_results["filter"]["predictive"].copy()

cols = ["t", 'with_monitor', 'what_detected', 'y', 'f', 'q', 'ci_lower',
        'ci_upper']
df_monitor["with_monitor"] = True
df_no_monitor["with_monitor"] = False
df_predictive = pd.concat(
  [df_monitor[cols], df_no_monitor[cols]]).reset_index(drop=True)

# Posterior
df_monitor_posterior = monitor_results["filter"]["posterior"].copy()
df_no_monitor_posterior = no_monitor_results["filter"]["posterior"].copy()

cols = ["t", 'with_monitor', 'parameter', 'mean', 'ci_lower', 'ci_upper']
df_monitor_posterior["with_monitor"] = True
df_no_monitor_posterior["with_monitor"] = False
df_posterior = pd.concat(
  [df_monitor_posterior[cols],
   df_no_monitor_posterior[cols]]).reset_index(drop=True)
# Join to get the data
df_posterior = df_posterior.merge(
  r.data_sim, on="t", how="left", validate="many_to_one")
```

---
# Predictive

```{r predictive-simulated, warning=FALSE, message=FALSE}
data_predictive <- py$df_predictive
with_detection <- data_predictive |> 
  dplyr::filter(with_monitor, what_detected != "nothing") |> 
  dplyr::pull(t)

data_predictive <- data_predictive |> 
  dplyr::mutate(which_shape = what_detected != "nothing" & what_detected != "",
                which_shape =  t %in% with_detection)

ggplot(data = dplyr::filter(data_predictive, t > 5), aes(x = t, y = y)) +
  geom_point(data = dplyr::filter(data_predictive, t > 5, with_monitor),
             aes(shape = which_shape), size = 1.6,  show.legend = FALSE) +
  geom_line(aes(y = f, col = with_monitor), show.legend = FALSE, size = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = with_monitor),
              alpha = 0.2) +
  scale_shape_manual(values = c(19, 4)) + 
  scale_color_manual(values = c(red, blue)) + 
  scale_fill_manual(values = c(red, blue)) + 
  geom_vline(xintercept = c(40, 80), linetype = "dotted", size = 1.4,
             col = "red") +
  scale_y_continuous(breaks = scales::pretty_breaks(8)) +
  scale_x_continuous(breaks = scales::pretty_breaks(10)) +
  labs(col = "With monitor?", fill = "With monitor?")

data_predictive |> 
  dplyr::filter(t > 10) |> 
  dplyr::group_by(with_monitor) |> 
  dplyr::summarise(MAPE = 100*mean(abs((y - f) / y)),
                   RMSE = sqrt(mean( (y - f)^2)), .groups = "drop") |> 
  dplyr::rename(Monitor = with_monitor) |> 
  kableExtra::kbl(digits = 2, caption = "One-step-ahead forecast metrics") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 12,
                            full_width = FALSE)
```

---
# Level

```{r level-simulated, warning=FALSE, message=FALSE}
py$df_posterior |> 
  dplyr::filter(parameter == "Intercept", t > 5) |>
  ggplot(aes(x = t, y = y)) +
  geom_point() +
  geom_line(aes(y = mean, col = with_monitor), show.legend = FALSE, size = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = with_monitor),
              alpha = 0.2) +
  scale_color_manual(values = c(red, blue)) + 
  scale_fill_manual(values = c(red, blue)) + 
  geom_vline(xintercept = c(40, 80), linetype = "dotted", size = 1.4,
             col = "red") +
  scale_y_continuous(breaks = scales::pretty_breaks(8)) +
  scale_x_continuous(breaks = scales::pretty_breaks(10)) +
  labs(col = "With monitor?", fill = "With monitor?")
```

---
# Growth

```{r growth-simulated, warning=FALSE, message=FALSE}
py$df_posterior |> 
  dplyr::filter(parameter == "Local Slope", t > 5) |>
  ggplot(aes(x = t, y = mean, col = with_monitor)) +
  geom_line(show.legend = FALSE, size = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = with_monitor),
              alpha = 0.2) +
  scale_color_manual(values = c(red, blue)) + 
  scale_fill_manual(values = c(red, blue)) + 
  geom_vline(xintercept = c(40, 80), linetype = "dotted", size = 1.4,
             col = "red") +
  scale_y_continuous(breaks = scales::pretty_breaks(8)) +
  scale_x_continuous(breaks = scales::pretty_breaks(10)) +
  labs(col = "With monitor?", fill = "With monitor?")
```


---
# Stored energy

- Monthly stored energy, measured in Gigawatt hours (GWh), in the Southeast
and Midwest regions of Brazil.

```{r stored-energy-viz, fig.width=16, fig.height=7}
py$data_ons |> 
  dplyr::mutate(time = as.Date(time)) |> 
  ggplot(aes(x = time, y = ear)) +
  geom_point(size = 2) +
  geom_vline(xintercept = as.Date(c("2012-02-01", "2016-02-01")),
             linetype = "dotted", size = 1.4, col = "red") +
  labs(x = "Time", y = "Stored energy (GWh)") +
  scale_x_date(breaks = scales::pretty_breaks(10), date_labels = "%b/%Y")
```

---
# The linear growth level model with seasonality

To reflect the month seasonality a two Fourier seasonal component of period 
$12$ is used. The regression vector and the evolution matrix are defined by

$$\mathbf{F}^\prime_t = (1, 0, 1, 0, 1, 0) \quad \textrm{and} \quad \mathbf{G}_t^{\prime} = \textrm{blockdiag}[\mathbf{G}_{T}, \mathbf{H}_{1}, \mathbf{H}_{2}]$$
where $\mathbf{G}_{T} = \begin{pmatrix} 1 & 1  \\ 0 & 1  \end{pmatrix}$ and

$$\mathbf{H}_j = \begin{pmatrix}  \cos(2\pi j / 12) & \sin(2\pi j / 12) \\ -\sin(2\pi j / 12) & \cos(2\pi j / 12)  \end{pmatrix}, \quad j = 1,2.$$

### Discount factors:
- $\delta_T = (0.95, 0.95) \rightarrow \delta^{\prime}_T = (0.30, 0.95)$.

- $\delta_S = 0.98$.


---
# Using the `Monitoring` class

```{python fit-ear, echo=TRUE}
a = np.array([data_ons["ear"].values[0:4].mean(), 0, 0, 0, 0, 0])
R = np.eye(6)
np.fill_diagonal(R, val=100)
mod = dlm(a0=a, R0=R, ntrend=2, deltrend=[0.95, 0.95], delseas=0.99,
          seasPeriods=[12], seasHarmComponents=[[1, 2]])
monitor = Monitoring(mod=mod)
fit_monitor = monitor.fit(
    y=data_ons["ear"], bilateral=True, prior_length=40, h=4, tau=0.135,
    discount_factors={"trend": [0.30, 0.95], "seasonal": 0.99})
```


```{python fit-ear-results, include=FALSE}
fit_no_monitor = monitor.fit(y=data_ons["ear"], bilateral=True,
                             prior_length=40, h=100000)
cols_chosen = [
    't', 'monitor',  'f', 'q', 'ci_lower', 'ci_upper', 'e', 'df',
    'H_lower', 'L_lower', 'l_lower', 'H_upper', 'L_upper', 'l_upper',
    'what_detected']

# Without monitor
data_predictive_no_monitor = fit_no_monitor.get(
    "filter").get("predictive").copy()

# With monitor
data_predictive_monitor = fit_monitor.get("filter").get("predictive").copy()

# Appending results
data_predictive_no_monitor["monitor"] = False
data_predictive_monitor["monitor"] = True
data_predictive = pd.concat(
    [data_predictive_no_monitor[cols_chosen],
     data_predictive_monitor[cols_chosen]])
data_predictive = data_predictive.merge(
    data_ons, on="t", how="left", validate="many_to_one")

# Smooth Posterior
data_posterior_no_monitor = fit_no_monitor.get(
  "smooth").get("posterior").copy()
data_posterior_no_monitor["monitor"] = False
data_posterior_monitor = fit_monitor.get("smooth").get("posterior").copy()
data_posterior_monitor["monitor"] = True
data_posterior = pd.concat(
    [data_posterior_no_monitor, data_posterior_monitor])
data_posterior = data_posterior.merge(
    data_ons, on="t", how="left", validate="many_to_one")

```


---
# Predictive

```{r predictive-ons-2, warning=FALSE, message=FALSE, fig.width=16, fig.height=8}
data_predictive <- py$data_predictive
with_detection <- data_predictive |> 
  dplyr::filter(monitor, what_detected != "nothing") |> 
  dplyr::pull(t)

data_predictive <- data_predictive |> 
  dplyr::mutate(which_shape = what_detected != "nothing" & what_detected != "",
                which_shape =  t %in% with_detection,
                time = as.Date(time))

ggplot(data = dplyr::filter(data_predictive, t > 5), aes(x = time, y = ear)) +
  geom_point(data = dplyr::filter(data_predictive, t > 5),
             aes(shape = which_shape), size = 1.6,  show.legend = FALSE) +
  facet_wrap(~monitor, ncol = 1) +
  geom_line(aes(y = f, col = monitor), show.legend = FALSE, size = 1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = monitor),
              alpha = 0.2) +
  scale_shape_manual(values = c(19, 4)) + 
  scale_color_manual(values = c(red, blue)) + 
  scale_fill_manual(values = c(red, blue)) + 
  geom_vline(xintercept = as.Date(c("2012-02-01", "2016-02-01")),
             linetype = "dotted", size = 1.4, col = "red") +
  scale_y_continuous(breaks = scales::pretty_breaks(6)) +
  scale_x_date(breaks = scales::pretty_breaks(10), date_labels = "%b/%Y") +
  labs(x = "Time", y = "Stored energy (GWh)", col = "With monitor?",
       fill = "With monitor?")

data_predictive |> 
  dplyr::filter(t > 10) |> 
  dplyr::group_by(monitor) |> 
  dplyr::summarise(MAPE = 100*mean(abs((ear - f) / ear)),
                   RMSE = sqrt(mean( (ear - f)^2))) |>
  dplyr::rename(Monitor = monitor) |> 
  kableExtra::kbl(digits = 2, caption = "One-step-ahead forecast metrics") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 12,
                            full_width = FALSE)
```



